\myTitle

### 2015-03-12 tex/response_to_refs_letter.tex

\documentclass[11pt]{letter}
\usepackage{graphicx}
\usepackage{color}

\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = 0.0 in
\evensidemargin = 0.0 in
\topmargin = 0 in
\headheight = 0.0 in
\headsep = 0.0 in
\parskip = 0.2in
\parindent = 0.0in

\date{}

\address{\mbox{}\vspace*{-1.0in}
\mbox{}\hfill
\includegraphics*{/Users/eriq/Documents/work/nonprj/NMFS_FORMS/NMFS_letterhead.pdf}\\}

\signature{Eric C. Anderson \\ \mbox{} \\ eric.anderson@noaa.gov\\ ph.~831-420-3983}

\newcommand{\reply}[1]{\begin{quotation}\small\sl\textcolor{blue}{#1}\end{quotation}}

\begin{document}

\begin{letter}{ \today

Dr. Nolan Kane\\
News and Views Editor\\
Molecular Ecology Resources \\
}

\opening{Dear Dr.~Kane:}

\reply{Some notes:  using flockture was mandatory because there is no other way to do a reproducible anaysis
with {\sc flock}.  Until people start distributing scriptable, reproducible tools, that is what we 
had to do.}

{\bf Editor Comments to Author:}

This paper, once revised, has the potential to shed light on some of the key similarities and differences among STRUCTURE and FLOCK, both of which are useful and important programs for population genetic analysis.  In that, I agree heartily with reviewer 2. However, reviewer 1 raises many important points that must be dealt with if this is to become acceptable. In particular, it would be important to broaden the scope of the article beyond a single dataset.  There are numerous well-analyzed datasets available these days.  For instance, take:

Recommendations for utilizing and reporting population genetic analyses: the reproducibility of genetic clustering using the program STRUCTURE
KJ Gilbert, RL Andrew, DG Bock, ...
Molecular Ecology 21 (20), 4925-4930

for some approaches that we used to assess repeatability of analytical procedures in quite a number of Structure datasets.  It would not be necessary to go quite as in-depth as in that article, but I do agree with reviewer 1 that a single dataset is insufficient.  Additionally it would be very helpful to use either simulated datasets or other datasets where we have a good idea of what the 'true' answer should be (e.g. some independent prior knowledge of some kind, or at the least a very thorough existing analysis to give reasonable expectations).

Reviewer 1 raises numerous other important points which will improve this paper substantially when dealt with, as does Reviewer 2 (though this reviewer's concerns tend to be much more minor). Despite the large number of suggestions between the reviewers, is should be quite possible to make these changes relatively easily - with the availability of so many existing, well-analyzed datasets it should be relatively easy to run at least a few more through FLOCK and STRUCTURE for a broader set of comparisons.  Alternately, it might make sense to do a fuller analysis of many datasets and re-submit as a full paper. Either would be quite useful to the Molecular Ecology and MER readership.

I look forward to a revised manuscript.

Best regards,
Dr. Nolan Kane
News and Views Editor, Molecular Ecology Resources

{\bf Referee Number 2 Comments:}

Editorial suggestions on Manuscript ID: MER-14-0343 ``Interpreting the FLOCK algorithm from a statistical perspective''
Page and Line numbers are from the pdf file.
\begin{itemize}

\item Page 2, Abstract. Line 18. See ``...no-admxixture...'' Spelling
\reply{fixed}
\item P. 2, Line 19. ``uncorrelated allele frequency prior albeit with more variability...''
\reply{done}
\item P. 3, Line 24. ``The best known example of this class ...''
\reply{done}
\item P. 3, Lines 44-45. ``Being able to interpret Recognizing FLOCK's in this fashion underpinnings should help users ... when it might and might not give substantially...''
\reply{superb edit. done}
\item P. 5. Line 68. ``...to randomly drawn values from the prior, ...''
\reply{done}
\item P. 5, Lines 70-71. ``...at iteration t = 0, 1, 2, ..., T,
1. each ...
\reply{done}
\item P. 6, Line 74. ``for all k = 1,..., K, implying the belief that the subpopulations are equally
represented in the sample before observing the genotypes present.
\reply{great. done.}
\item Page 7, Lines 107-108. ``...the program randomly allocates or each of partitions the N
individuals into (approximately) one of K equal-sized clusters.
\reply{done}
\item Page 7, Lines 109- ...each individual is given the chance to be reallocated (i.e., moved to a different cluster). This is done on the basis of maximum likelihood: non-randomly assigned to the cluster for which its multilocus likelihood is maximal: as Duchesne \& Turgeon say ...''
\reply{This is a great edit. we've incorporated it.}
\item Page 8, Line 138. ``If, all that was desired...'' Delete comma. 
\reply{done}
\item Page 9, Line 155. ``... we can expect that it may be more is susceptible to ...''
\reply{this line got edited out in other changes}
\item Page 9, Line158. ``...the same solution among repeat applications to the same data set less consistently than does STRUCTURE.''
\reply{This line was removed/changed in a previous edit}
\item Page 10, Line 161. ``...to compare the programs FLOCK and STRUCTURE.'' [add] The criteria for comparison include amount of computer time used, estimates of K, and the allocation of individuals among clusters.'' Now move Lines 177 -- 182 to immediately follow, i.e., ``For FLOCK, the user must specify ... is calculated for each individual.'' Continue with Lines 161 -168, i.e., ``FLOCK was run six times ... running 64-bit Windows 7.'' You might also want to explain the choice of 50,000 sweep burn-in and 150,000 sweep sample.
\reply{These lines were completely overhauled given the changes in our simulation assessment}
\item Page 10. Lines 169 `` 175. No change except Page 10, Line 175. ``DISTRICT'' spelling
\reply{done}
\item Page 10, Lines 176 `` 198. Omit Lines 177- 182 that have been moved above. Page 11, Line 197. ``...for estimating K with FLOCK to results obtained from
STRUCTURE using ln P(D) ...''
\reply{eliminated already}
\item Page 12, Lines 209 -- 228. Figure 1 is referenced which uses lower case k, while this text
uses the upper case K.
\reply{eliminated already}
\item Page 12, Line 217. ``Both programs idenfity ...'' Spelling
\reply{done}
\item Page 12, Lines 230 `` 233. ``To investigate the stability of the partition arrangement of individuals within from the reference populations to move around the partition space, we varied ... Unfortuantely Spelling only the ...is given as FLOCK output.''
\reply{done}
\item Page 13, Line 234. ``...during the last later iterations of runs, individuals ...''
\reply{eliminated already}
\item Page 14, Lines 247--248. ``...give similar results, only although at values of K > 4, ...
appear to be more variable among random starting partitions''.
\reply{eliminated already}
\item Page 14, Lines 260--262. ``This, coupled with ...of length zero, complicates ... from
FLOCK. if no non zero plateaus are encountered.
\reply{eliminated already}
\item Page 15, Lines 270--271. `` ... in their comparisons, we did not observe this advantage,
probably for in our analysis. This appears to result from two issues reasons. ``
\reply{the whole discussion of speed was completely revamped once we implemented {\sc flockture}.}
\item Page 16, Lines 302--303. ``This is especially true in real populations and the actual conditions under which samples are drawn which may not conform to the assumptions of genetic clustering models.''
\reply{eliminated already}
\item Page 16, Lines 304--307. ``We made our comparison between FLOCK and the no- admixture model in STRUCTURE which is not the instead of its default with-admixture model. used in STRUCTURE. Rather, by default, STRUCTURE uses its with admixture model. Such a The default model explicitly tries to account for the admixed origin of individuals in the sample and thus may be is more appropriate model than the no admixture model, or FLOCK, when the sample contains such individuals. that are admixed between populations.''
\reply{eliminated already}
\item Page 17, Line 318. ``... with correlated allele frequencies seems like it would seem an interesting challenge, ...''
\reply{eliminated already}
\item Page 22 List of Figures. Lines 386 - 393. Program DISTRUCT plots of cluster membership probabilities of steelhead individuals (qi values from STRUCTURE and the normalized likelihood values from FLOCK). Individual fish are grouped left to right into their geographic samples, and each horizontal plot represents a run of the program
STRUCTURE, or the ``best run'' from a series of runs in FLOCK. Within horizontal plots, each steelhead individual is represented by a vertical bar composed of colored segments of lengths proportional to the various cluster membership probabilities. The numbers in parentheses ... same visually indistinguishable ``best run.''
\reply{This is an excellent, concise description of those plots. We thank the referee for 
providing it, and we have used it.}
\end{itemize}

{\bf Referee Number 1 comments}

I have several important concerns over this paper:

\reply{I'll give brief replies to each of the following here as an outline of how we
changed the paper, and will give more complete details in the following sections}
\begin{enumerate}
\item The conceptual framework is inappropriate since FLOCK is not an MCMC algorithm
\reply{It is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
recognizing it as a special case of well known statistical methods is appropriate.}
\item There is only one dataset which is way too little to come to conclusions
\reply{We agree with the referee. We have undertaken a simulation study across a range of different $F_\mathrm{ST}$ values.}
\item There are basic flaws in the interpretation of the outputs of FLOCK, leading to invalid
comparisons relative to K values other than the estimate
\reply{We have eliminated issues of what can and can't be said about allocation of individuals  at a
certain value of $K$ by doing all the simulations and the analyses at a known value of $K$.}
\item Comparisons of estimates of K cannot be performed with sufficient rigor
\reply{We agree with the referee here.  In fact, we weren't entirely comfortably addressing the issue of estimating
$K$ in the previous version of the manuscript, lest it appear that we were implicitly encouraging it. 
Therefore, we now focus entirely on allocation of individuals.}
\item The speed comparisons are based on an ill-chosen processing unit
\reply{We implemented the {\sc flock} algorithm by modifying the C source code for {\sc structure} and
adding an R wrapper.  This allows a more direct comparison of the amount of computation involved.  If 
{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  But since it
runs 200 times slower than it should for what it is doing, there doesn't seem much speed advantage in 
the data sets we ran.  We report that now in the paper in a way that should be easily understood. }
\end{enumerate}

I now describe each of those concerns in more detail.

{\bf 1) Conceptual framework}

It must be emphasized that FLOCK is not an MCMC based algorithm. It has no target function, no transition probabilities and does not move in small measured probabilistic steps. No amount of mathematical analysis will ever change this.

\reply{I don't believe that we said {\sc flock} was an MCMC algorithm.  We noted that it was a limiting
 case of a simulated annealing algorithm.}

In fact, similar iterative methods have been around for a very long time (e.g. Newton's method for finding roots), much before the first MCMC algorithms were designed (1950?s). This family of methods should be analysed within a truly pertinent conceptual framework (attractors, basins of attraction, fixed points, orbits, etc). The book A First Course in Discrete Dynamical Systems (RA Holmgren, 1994) provides an excellent introduction to this topic. 
\reply{The referee is correct that iterative methods have been around for a long time.
In fact, rather than viewing it as special case of simulated annealing, one could just as well
note that {\sc flock} is a type of gradient descent-algorithm attempting to maximize the
likelihood of the data given a configuration of clusters.
Regardless of what name is given to it, since it is effectively an algorithm
that (intentionally or not) yields solutions with 
high (and possibly maximal) likelihood, we have difficulty understanding
how the referee can honestly declare that, for purposes of analysis of the algorithm,
the conceptual framework of dynamical systems is any more ``truly pertinent'' than
the conceptual framework of statistics.   }

The current popularity of MCMC methods in the field of genetic analysis does not justify trying to cast every current algorithm into an MCMC framework.

\reply{Again, we are not so much casting it in an MCMC framework as we are in a 
statistical/probabilistic framework.  Since it is a problem in statistical inference
it seems to us reasonable to understand it in terms of the probabilities involved. Those
probabilities are, for the most part, identical to those in the program {\sc structure} because
the underlying model is the same in the two programs.  Regarding casting the algorithm in
comparison to {\sc structure}, it makes sense to us to try to understand and explain the {\sc flock}
algorithm in those terms since {\sc structure} is so well known and because the models
underlying it and {\sc flock} are essentially identical.  }

For example:
\begin{itemize}
\item Line 156: it is suggested that Flock can be trapped in a local mode, this is incorrect because Flock does NOT search for maximum values, either local or else
\reply{We thank the referee for pointing out our erroneous language here.  We should not have used
the word ``trapped.''  We have updated that sentence to read, ``As such, we should not be surprised
if it is prone 
to converging to different parts of the posterior probability surface on runs started
from different initial conditions.}

\item Line 254: it is suggested that Flock is searching over all possible partitions. Flock does not sample nor search the surface or space of all partitions.
\reply{We would not have wanted it to sound like we thought {\sc flock} was searching
each and every possible
partition, since this would be infeasible in most problems, so we thank the referee for pointing out that our reference to a space of all partitions might
have confused readers in this way.  We have changed that section appropriately.  However, on whether
or not {\sc flock} samples from the space of partitions, we respectfully submit that the referee is
incorrect when he states that, ``Flock does not sample nor search the surface or space of all partitions.'' 
Consider this: 1) for each run, the initial starting partition in {\sc flock} is chosen (or you could say,
sampled) randomly from amongst
possible partitions with roughly the same number of individuals; 2) the starting partition uniquely
determines the partition that the algorithm will converge to since it is deterministic, 3) different
starting partitions can lead to different final solutions/partitions.  How is it then that each
run in {\sc flock} is {\em not} a sample from the partitions?  }
\item Line 234: ?individuals would alternate between reference groups?: this, precisely, is an orbit
\reply{We removed discussions of individuals alternating between reference groups.}
\end{itemize}

{\bf 2) Data}

Performance comparisons between programs based on a single dataset are of little value. The fact that it is large and real does not compensate. For this purpose, one would expect that a large set of simulations spanning several parameter values with or without several real datasets would be used instead of a single empirical dataset. The advantage of simulations is, of course, that the true genetic structure of the data (essentially K) is known and so the relative performances may be compared on a solid basis. Here, unless I missed something, there is no independent (of the two programs) source of knowing with a reasonable degree of certainty the actual number of clusters represented within this collection of genotypes. In fact, there appears to be several candidate values for K with this particular dataset (see below).

\reply{We agree whole-heartedly and undertake a numerous simulations of data like the real
steelhead data in the previous submission.  Having done this, and analyzed the results,
we can now address the comments below.}
Several (over)generalizations are based on this unique dataset:
\begin{itemize}
\item (p.9 ) FLOCK seems to find the same solution less consistently than does STRUCTURE
\reply{From our simulations using {\sc flockture} this appears to be true.  One need only look
at the variability present in the purple boxplots in Figures~2 and~3 to see that there is more
variability between {\sc flockture} analyses (50 runs of 20 iterations each) on the same data set 
than between {\sc structure} analyses. }
\item (p.14) The problem of getting caught in local modes is not unique to FLOCK...; however our results suggest it is a bigger problem for FLOCK
\reply{The simulation results bear this out, but we have reworded it to, ``This likely 
increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions in 
each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.''
}
\item (p.16) Given FLOCK'S tendency to converge to different solutions with large K and n...
\reply{See above.  Note that we don't vary $K$ now so we don't call it a problem ``with large $K$'' any
longer.}
\item (p.16) It is probably better to use STRUCTURE in such cases...
\reply{We don't say this, but we do point out that {\sc flock} is not best-suited to high-gene-flow
species with very low $F_\mathrm{ST}$.}
\end{itemize}

{\bf 3) Interpretation of FLOCK outputs (purpose of FLOCK, best run, stochasticity of plateau lengths)
Purpose of FLOCK}

I agree with the following statement (p. 11):
``While we take the stance that estimates of K made with any unsupervised clustering algorithm from data on real (non-idealized) populations should always be interpreted and used cautiously...''
I am fully aware that we are dealing with probabilistic statements here, not certainties. However, I am also convinced that the search for the number of genetic units (K) is a central issue. This is precisely what Flock intends to do, and it is also the main task that STRUCTURE was built for.

\reply{We disagree with the referee that estimating $K$ was the ``main task that {\sc structure} was
built for.''  One of us had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
{\sc structure} paper was published. At that time, the members of the Donnelly lab were not eagerly using
{\sc structure} to estimate $K$ from different samples; they were using {\sc structure} to correct
for population admixture when doing association tests in human genetics---our understanding
is {\em that} was the task that structure was original built for.  Furthermore, when I contacted
Jonathan a few years later to ask him some details about how $Ln P(D)$ was calculated by {\sc structure},
he pointed out to me (while getting his notes out of his filing cabinet) that the method for estimating
$K$ was added into the paper to satisfy a referee who had requested it, and that Matthew Stephens 
(one of the co-authors) was not ever extremely pleased to have it in there.  Finally, I teach an annual
course with Matthew Stephens, and each year he explains to the students why estimating $K$ is so
difficult and why estimates of $K$ from {\sc structure} or any other procedure should be 
regarded with circumspection.  }

Thus, the basic purpose of FLOCK is to estimate K and if the estimate is not undecided, to obtain a reasonably accurate partition of the collection of genotypes. Consistency across different values of K is not intended. By considering outputs for any value of K, the authors make invalid comparisons. Comparisons among clustering solutions for different values of K provided by STRUCTURE are commonly reported in the literature, and the successive clustering solutions may at times make biological sense. That practice, however, is not recommended in FLOCK and therefore cannot serve as basis for evaluating its results outside the estimated value of K.

### 2015-03-12 tex/main-body-text.tex
Addressing reviewer comments.

\section*{Introduction}
Methods for the unsupervised clustering of genotypes have received 
considerable attention in the molecular ecology literature.  
The best known example of this class of methods is {\sc structure} 
\citep{Pritchardetal2000}, with over 12,000 citations identified on 
Google Scholar.  A variety of other 
clustering methods have been developed that are all closely related to 
{\sc structure}, for example {\sc NewHybrids} \citep{And&Tho2002}, {\sc 
BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
\citep{Coranderetal2004}. An overview of the similarities between several of these methods can be 
found in \citet{Anderson2009PGAC}.

@@

Being able to interpret {\sc flock} in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration,
and to aid in application to simualted data, we
create the progam {\sc flockture} by adapting the code of {\sc structure} to 
use a simulated annealing approach. We then compare results of the two
programs from large 
datasets simulated with varying degrees of population differentiation.

@@

After initializing the unknown allele frequencies, $\theta = (\theta_1,\ldots,\theta_K)$, to randomly drawn values,
$\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
$Z_i$'s and $\theta$ using Gibbs sampling.  That is,
at iteration $t = 0, 1, 2, \ldots$:
\begin{enumerate}
\item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
the full conditional distribution of $Z_i$ given
$Y_i$ and $\theta^{(t)}$ (the current estimate of the allele frequencies).  This
distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
\item $\theta^{(t)}$ is updated to $\theta^{(t+1)}$ from its full conditional distribution.  For each 
cluster
$k$ and locus $\ell$, the full conditional for $\theta_{k\ell}$ is independently a 
Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
$j=1,\ldots, A_\ell$,
where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
individuals with $Z^{(t+1)}_i = k$.   
\end{enumerate}

@@

The first step of the {\sc flock} algorithm is initialization, during which 
the program randomly allocates each of the $N$ individuals to one of 
$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is given the chance to be reallocated 
(\ie moved to a different cluster).  This is done on the basis of 
maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
performed following multilocus maximum likelihood (Paetkau et al. 1995).''

@@

It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
algorithm for the {\sc structure} model in which the cooling schedule is
``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
the algorithm.''  As such, we should not be surprised if it is prone 
to becoming trapped in local modes of the posterior probability, though
the practical implications of that are not yet clear.

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
 runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became somewhat inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence and also
the large size of the search space over possible partitions. 
The likelihood surface defined on partitions of this space will often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions. The problem of getting 
caught in local modes is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulationssuggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each run of {\sc flock}.
While \citet{Duc&Tur2012} observed faster processing times (relative to other methods), 
their comparision was based on setting {\sc flock} to perform 50 runs and 20 iterations for each 
$K$ starting at 2 and ending when one of the stopping conditions was met, and computing times for that analysis 
were compared to ten iterations of the admixture and correlated allele frequency model in {\sc structure} using 
a 50,000 sweep burn-in followed by a 200,000 sweep sample 
from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. It is difficult to reliably assess differences in program run times when there is no clear consensus on what 
constitutes a reasonable comparison with respect to run length and number of runs, however, we show that if 
{\sc flock} were to be implemented using good coding practices in a fast language, rather than as
a VBA routine in Microsoft Excel,
it would be approximately 200 times faster. This is not an inconsequential difference and is one reason
we used {\sc flockture} for a large-scale simulation assessment of the {\sc flock} algorithm.

>>>>

\section*{Introduction}
Methods for the unsupervised clustering of genotypes have received 
considerable attention in the molecular ecology literature.  
The best known of this class of methods is {\sc structure} 
\citep{Pritchardetal2000}, with over 12,000 citations identified on 
Google Scholar.  A variety of other 
clustering methods have been developed that are all closely related to 
{\sc structure}, for example {\sc NewHybrids} \citep{And&Tho2002}, {\sc 
BayesAss+} \citep{Wil&Ran2003}, and {\sc baps} 
\citep{Coranderetal2004}. An overview of the similarities between several of these methods can be 
found in \citet{Anderson2009PGAC}.

@@

Recognizing {\sc flock}'s underpinnings in this fashion should help users
interpret output from {\sc flock} and predict when it might and might not
give substantially different results than {\sc structure}. Below we describe
the correspondence between the programs in more detail, and then, for illustration,
and to aid in application to simualted data, we
create the progam {\sc flockture} by adapting the code of {\sc structure} to 
use a simulated annealing approach. We then compare results of the two
programs from large 
datasets simulated with varying degrees of population differentiation.

@@

After initializing the unknown allele frequencies, $\theta = (\theta_1,\ldots,\theta_K)$, to randomly drawn values from the prior,
$\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
$Z_i$'s and $\theta$ using Gibbs sampling.  That is, if Gibbs sampling is run for $T$ iterations,
at iteration $t =0, 1, 2, \ldots, T-1$:
\begin{itemize}
\item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
the full conditional distribution of $Z_i$ given
$Y_i$ and $\theta^{(t)}$ (the current estimate of the allele frequencies).  This
distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$, implying the belief that the subpopulations are equally
represented in the sample before observing the genotypes present.    
\item $\theta^{(t)}$ is updated to $\theta^{(t+1)}$ from its full conditional distribution.  For each 
cluster
$k$ and locus $\ell$, the full conditional for $\theta_{k\ell}$ is independently a 
Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
$j=1,\ldots, A_\ell$,
where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
individuals with $Z^{(t+1)}_i = k$.   
\end{itemize}

@@

The first step of the {\sc flock} algorithm is initialization, during which 
the program randomly allocates or partitions the $N$ individuals into
approximately equal-sized
clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is non-randomly assigned to the cluster for which its multilocus likelihood is maximal: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
performed following multilocus maximum likelihood (Paetkau et al. 1995).''

@@

It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
algorithm for the {\sc structure} model in which the cooling schedule is
``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
the algorithm.''  As such, we should not be surprised if it is prone 
to converging to different parts of the posterior probability surface on runs started
from different initial conditions.

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{
  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
  Individuals are grouped left to right into their simulated population samples, and each 
  horizontal plot represents a run of the program
  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
  Within horizontal plots, each individual is represented by a vertical bar composed of colored
  segments of lengths proportional to the various cluster membership probabilities.
  Each horizontal plot
  represents one of nine 
 runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
parentheses following the model names give the number, out of 9, of replicate runs
that obtained the same visually-indistinguishable estimate. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each execution of {\sc flock}.
\citet{Duc&Tur2012} observed faster processing times relative to {\sc structure}.
Having implemented {\sc flockture} it is clear that the amount of computation required for a
single sweep of {\sc structure} is equivalent to an iteration of reallocation of individuals
in {\sc flock}.
Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
computations, one would expect that running flock once with 50 runs, each of 20 iterations,
would require about the same amount of time as running {\sc structure} on the same
data set for 1,000 sweeps.  Since it is recommended to run {\sc structure} longer than 1,000 sweeps, {\sc flock} should have a speed advantage. However, {\sc flock} requires far more 
time for each computation
than {\sc structure}, as is clear from the fact that it runs 200 times slower than 
{\sc flockture}.  Thus, for data sets comparable to the ones we simulated, one can expect
50 runs of 20 iterations in {\sc flock} (a standard execution of the program) to require as
much time as running {\sc structure} for 200,000 sweeps. This does not constitute
a speed advantage.
If {\sc flock} were to be implemented using good coding practices in a fast language,
rather than as a VBA routine in Microsoft Excel,
it could be made much faster. As it is, we found it preferable to implement our own version
in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.
